---
title: "R Notebook"
output: html_notebook

#it takes a while to runn this part, so we have provided you with direct output file
#Xp_hierbaps.csv is present in the github respository, please download that file before running this

---
```{r}
# Installs packages for managing sequence data, computing population structures, and basic plotting
install.packages(c("ape", "rhierbaps", "ggplot2", "ggnewscale"))
# 1. Install the Bioconductor manager first
install.packages("BiocManager")
install.packages("phangorn")   # only needed once


# 2. Use BiocManager to download and install the specialized tree layout packages
BiocManager::install(c("ggtree", "ggtreeExtra"))
```


```{r}
# Load sequence and clustering tools
library(ape)
library(rhierbaps)

# Load plotting and tree visualization tools
library(ggplot2)
library(ggnewscale)
library(ggtree)
library(ggtreeExtra)
library(phangorn)
```


#because size of snp files was big 
dna <- read.dna(
  "/Users/azk0151/Library/CloudStorage/Box-Box/Amanpreet/APS_2026_workshop/Result_files/Parsnp/SNPs.fa",
  format = "fasta"
)

seg <- seg.sites(dna)

dna.snp <- dna[, seg]

dim(dna.snp)

write.dna(
  dna.snp,
  file = "/Users/azk0151/Library/CloudStorage/Box-Box/Amanpreet/APS_2026_workshop/Result_files/Parsnp/SNPs_variable_only.fa",
  format = "fasta",
  nbcol = -1,
  colsep = ""



#now use this reduced version
snp.matrix <- load_fasta("/Users/azk0151/Library/CloudStorage/Box-Box/Amanpreet/APS_2026_workshop/Result_files/Parsnp/SNPs_variable_only.fa")
#to determine if loaded correctly 
class(snp.matrix)
dim(snp.matrix)

#Run hierbaps
Perforans.hb <- hierBAPS(snp.matrix, max.depth=2, n.pops=20, quiet=FALSE)
#Inspect the result
head(Perforans.hb$partition.df)

#Save the results
#write.csv(Perforans.hb$partition.df, file = "/Users/azk0151/Downloads/Xp_hierbaps.csv",
    col.names = TRUE, row.names = FALSE)


#it takes a while to runn this part, so we have provided you with direct output file
#Xp_hierbaps.csv is present in the github respository, please download that file before running this

```{r}
## 1. Load the tree file

tree <- read.tree("~/Downloads/Part1B/parsnp.tree")


## 2. Root the tree using midpoint rooting

tree <- midpoint(tree)


## 3. Initialize the rooted rectangular tree
# Tip labels are shown with dotted lines connecting them to the branches

p <- ggtree(tree, layout = "rectangular") +
  geom_tiplab(
    size = 1,
    align = TRUE,
    color = "gray",
    linetype = "dotted",
    linesize = 0.3,
    offset = 0.05
  )


## 4. Prepare Heatmap Data

heatmapData <- read.csv(
  "~/Downloads/Part1B/Xp_level_1_heirbaps.csv",
  row.names = 1
)

rn <- rownames(heatmapData)

heatmapData <- as.data.frame(
  sapply(heatmapData, as.character)
)

rownames(heatmapData) <- rn


## 5. Define Colors

heatmap.colours <- c(
  "#ffffff",
  "orange",
  "#A6CEE3",
  "blue",
  "#CAB2D6",
  "#B2DF8A",
  "#CC0066",
  "#CC00CC",
  "#FFC0CB",
  "#8E8E38",
  "#006400"
)

names(heatmap.colours) <- 0:10


## 6. Combine Tree and Heatmap

p2 <- gheatmap(
  p,
  heatmapData,
  offset = 0.1,
  width = 0.05,
  font.size = 1.25,
  colnames_position = "top",
  colnames_angle = 360,
  colnames_offset_y = 0,
  hjust = 0
) +
  scale_fill_manual(
    name = "SCs",
    values = heatmap.colours,
    labels = c(
      "1", "2", "3", "4", "5",
      "6", "7", "8", "9", "10"
    )
  )


## 7. Display the tree

p2


## 8. Save the figure

ggsave(
  filename = "~/Downloads/ParSNP_tree_SCs.png",
  plot = p2,
  width = 12,
  height = 10,
  units = "in",
  dpi = 1200
)
```

