---
title: "R Notebook"
output: html_notebook

#it takes a while to runn this part, so we have provided you with direct output file
#Xp_hierbaps.csv is present in the github respository, please download that file before running this

---
```{r}
# Installs packages for managing sequence data, computing population structures, and basic plotting
install.packages(c("ape", "rhierbaps", "ggplot2", "ggnewscale"))
# 1. Install the Bioconductor manager first
install.packages("BiocManager")

# 2. Use BiocManager to download and install the specialized tree layout packages
BiocManager::install(c("ggtree", "ggtreeExtra"))
```


```{r}
# Load sequence and clustering tools
library(ape)
library(rhierbaps)

# Load plotting and tree visualization tools
library(ggplot2)
library(ggnewscale)
library(ggtree)
library(ggtreeExtra)
```


#because size of snp files was big 
dna <- read.dna(
  "/Users/azk0151/Library/CloudStorage/Box-Box/Amanpreet/APS_2026_workshop/Result_files/Parsnp/SNPs.fa",
  format = "fasta"
)

seg <- seg.sites(dna)

dna.snp <- dna[, seg]

dim(dna.snp)

write.dna(
  dna.snp,
  file = "/Users/azk0151/Library/CloudStorage/Box-Box/Amanpreet/APS_2026_workshop/Result_files/Parsnp/SNPs_variable_only.fa",
  format = "fasta",
  nbcol = -1,
  colsep = ""



#now use this reduced version
snp.matrix <- load_fasta("/Users/azk0151/Library/CloudStorage/Box-Box/Amanpreet/APS_2026_workshop/Result_files/Parsnp/SNPs_variable_only.fa")
#to determine if loaded correctly 
class(snp.matrix)
dim(snp.matrix)

#Run hierbaps
Perforans.hb <- hierBAPS(snp.matrix, max.depth=2, n.pops=20, quiet=FALSE)
#Inspect the result
head(Perforans.hb$partition.df)

#Save the results
#write.csv(Perforans.hb$partition.df, file = "/Users/azk0151/Downloads/Xp_hierbaps.csv",
    col.names = TRUE, row.names = FALSE)


#it takes a while to runn this part, so we have provided you with direct output file
#Xp_hierbaps.csv is present in the github respository, please download that file before running this

```{r}


## 1. Load the tree file # add the path here to your download directory where the Xp_hierbaps.csv  is located in your computer

tree <- read.tree("~/Downloads/Part1B/parsnp.tree")

## 2. Load the list of specific genomes to highlight
# Ensure this CSV has a column 'label' containing the names to highlight
highlight_list <- read.csv("~/Downloads/Part1B/highlights.csv")$label

## 3. Initialize the Circular Tree
# We set layout='circular' and use a logical check inside geom_tiplab 
# to show names only if they are in our 'highlight_list'
p <- ggtree(tree, layout='rectangular') + 
  geom_tiplab(aes(label = ifelse(label %in% highlight_list, 
                                 paste0("\u2794 ", label), "")), 
              size = 1, align = TRUE, color = "gray", 
              linetype = "dotted", linesize = 0.3, offset = 0.05) +
  # Add a small marker at the tip for highlighted ones
  geom_tippoint(aes(subset = label %in% highlight_list), 
                shape = 17, color = "darkred", size = 5)

## 4. Prepare Heatmap Data
heatmapData <- read.csv("~/Downloads/Part1B/Xp_level_1_heirbaps.csv", row.names = 1)
rn <- rownames(heatmapData)
heatmapData <- as.data.frame(sapply(heatmapData, as.character))
rownames(heatmapData) <- rn

## 5. Define Colors
heatmap.colours <- c("#ffffff", "orange", "#A6CEE3", "blue", "#CAB2D6","#B2DF8A","#CC0066","#CC00CC", "#FFC0CB", "#8E8E38","#006400")

names(heatmap.colours) <- 0:10

## 6. Combine Tree and Heatmap
# Note: You may need to adjust the 'offset' value based on the radius of your circular tree
p2 <- gheatmap(p, heatmapData, offset = 0.1, width = 0.05, 
               font.size = 1.25, colnames_position = "top", 
               colnames_angle = 360, colnames_offset_y = 0, hjust = 0) + 
      scale_fill_manual(name="SCs", values=heatmap.colours, 
                        labels=c("1", "2", "3", "4", "5", "6", "7", "8", "9", "10"))

p2

ggsave(filename = "~/Downloads/ParSNP_tree_SCs.png", plot = p2, width = 12, height = 10, units = "in", dpi = 1200)
```

